On the propriety of restricted Boltzmann machines

### 2016-11-11 writing/draft.Rmd
section 2

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common. In this paper, we consider the binary case of a RBM for concreteness. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no dependency connections within a layer. An example of this structure is in Figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.

\begin{figure}[ht]
  \centering
  \resizebox{\linewidth}{!}{\input{tikz/rbm.tikz}}
  \caption{An example restricted Boltzmann machine (RBM), consisting of two layers, one hidden ($\mathcal{H}$) and one visible ($\mathcal{V}$), with no connections within a layer. Hidden nodes are indicated by gray circles and the visible nodes indicated by white circles.}
  \label{fig:rbm}
\end{figure}

A common use for RBMs is to create features for use in classification. For example, binary images could be classified where the pixel values would be treated as the visible variables $v_i$ [@hinton2006fast].

### Joint distribution
Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in an RBM for some integers $\nv, \nh \ge 1$. Each single "binary" random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$ with "$1$" always indicating the "high" value of the variable. While $\mathcal{C}=\{0,1\}$ may be a natural or naive starting point, we show later that the coding $\mathcal{C}=\{-1,1\}$ induces better model properties for the RBM. We posit a parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes as

@@

where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)$ is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one. For 
\begin{align}
\boldsymbol t(\boldsymbol x) = (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}, \label{eqn:t}
\end{align}
let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\}$ be the vector of variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the neg-potential function. Note that the RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects. The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.

@@

RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast, @salakhutdinov2012efficient, or @srivastava2013modeling for examples]. By stacking multiple layers of RBMs in a deep architecture, proponents of the models claim to produce the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp. 450]. The stacking is achieved by treating a hidden layer of one RBM as the visible layer in a second RBM, and so on until the desired multi-layer architecture is created.

@@

### Near-degeneracy
In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph. Here, as earlier, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function. Let $C$ denote the convex hull spanned by the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$. @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$. Conceptually, this makes sense because if a model is near-degenerate, then there are only a small number of elements of the sample space $\mathcal{X}$ that have positive probability, meaning the the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is only averaging that same number of values of $t( \boldsymbol x)$. Thus, this average value will not be pulled away from the boundary of $C$.

The RBM class of models can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space, $\mathcal{C}^{\nv + \nh}$. Using the method of @handcock2003assessing, when random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$. Here the mean of $\boldsymbol t(\boldsymbol x)$ is
$$
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
$$

@@

Considering a class of exponential families of distributions, @schweinberger2011instability introduced a concept of model defficiency related to *instability*. Instability can be roughly characterized as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$. To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how the RBM model might be expanded to incorporate more and more visibles. When increasing the size of a RBM model, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used). To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $V$ of visibles $(V_1, \dots, V_V)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{V}$ as
\begin{align}
\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) \label{eqn:elpr}
\end{align}
where $P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{\{h_j\}\in \mathcal{C}}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_\nv)$ for the visible variables $(V_1, \dots, V_\nv)$ under parameters $\boldsymbol \theta_\nv$.

In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio (\ref{eqn:elpr}) needs to stay bounded for the sequence of RBM models to be stable. That is, we make the following convention

\begin{defn}[S-unstable RBM]
Let $\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh +  \nh*\nv}, \nv \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $\nh \equiv \nh(\nv) \ge 1$, can be an arbitrary function of the number $\nv$ of visibles. A RBM model formulation is \emph{Scheinberger-unstable} or \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) \equiv \lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta}(v_1, \dots, v_\nv)}\right] = \infty.
\end{align*}
\end{defn}
In other words, given any $C > 0$, there exists an integer $n_C > 0$ so that $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > C$ for all $\nv \ge n_C$.

@@

S-unstable RBM model sequences are undesirable for several reasons. One is that, as mentioned above, small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
$$
\Delta(\boldsymbol \theta_\nv) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{P_{\boldsymbol \theta_\nv}(v_1^*, \dots, v_\nv^*)} : (v_1, \dots, v_\nv) \& (v_1^*, \dots, v_\nv^*) \in \mathcal{C}^\nv \text{ differ by exactly one component}\right\},
$$
denoting the biggest log-probability ratio for a one-pixel (one component) change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_\nv$. We then have the following result.

@@

In other words, if the probability ratio (\ref{eqn:elpr}) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., instability).

Additionally, S-unstable RBM model sequences are connected to near-degeneracy of section \ref{near-degeneracy}, in which model sequences place all probability on a small portion of their sample spaces. To see this, define a modal set of possible outcomes (i.e. set of highest probability outcomes) in RBM models with parameters $\boldsymbol \theta_\nv, \nv \ge 1$ as
$$
M_{\epsilon, \boldsymbol \theta_\nv} \equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v) > (1-\epsilon)\max\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
$$
for a given $0 < \epsilon < 1$. Then S-unstable model sequences are guaranteed to be degenerate, as the following result shows.

@@

### Uninterpretability
For spatial Markov models, @kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by dependence parameters in a model being so extreme that marginal mean-structures fail to hold as anticipated by the model statement. We adapt this notion to RBM models as follows. Note that, in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_\nv}$ and $\theta_{h_1}, \dots, \theta_{h_\nh}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_\nv, H_1, \dots, H_\nh$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh$. Hence, these main effect parameters have a clear interpretation under an independence model (one with $\theta_{ij} = 0$) but this interpretation (as well as an understanding of the nature of dependence) can break down as interaction parameters increase in magnitude relative to the size of the main effects. In such case, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the modes (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (destroying the model interpretation of dependence as local conditional modifications of all overall marginal mean structure). To assess which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ given independence where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$. Hence, $\boldsymbol \theta^*$ and $\boldsymbol \theta$ have the same main effects but $\boldsymbol \theta^*$ has $0$ dependence parameters. Uninterpretability is then avoided at a parametric specification $\boldsymbol \theta$ if the model expected value at $\boldsymbol \theta$ is not very different from the corresponding model expectation under independence. Using this, it is possible to investigate what parametric conditions lead to uninterpretability in a model versus those that guarantee interpretable models. If $||\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]||$ is large, then the RBM model with parameter vector $\boldsymbol \theta$ is said to be uninterpetable. The quantities to compare in the RBM case are

>>>>

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being the case most commonly considered. In this paper, we consider the binary case of a RBM for concreteness. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no dependency connections within a layer. An example of this structure is in Figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.

\begin{figure}[ht]
  \centering
  \resizebox{\linewidth}{!}{\input{tikz/rbm.tikz}}
  \caption{An example restricted Boltzmann machine (RBM), consisting of two layers, one hidden ($\mathcal{H}$) and one visible ($\mathcal{V}$), with no connections within a layer. Hidden nodes are indicated by gray filled circles and the visible nodes indicated by unfilled circles.}
  \label{fig:rbm}
\end{figure}

A common use for RBMs is to create features for use in classification. For example, binary images can be classified where the pixel values are treated as the visible variables $v_i$ [@hinton2006fast].

### Joint distribution
Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in an RBM for some integers $\nv, \nh \ge 1$. Each single binary random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set, $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$, with "$1$" always indicating the "high" value of the variable. While $\mathcal{C}=\{0,1\}$ may be a natural or naive starting point, we argue later that the coding $\mathcal{C}=\{-1,1\}$ induces more interpretable model properties for the RBM. A standard parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes is

@@

where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)$ is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one. For 
\begin{align}
\boldsymbol x &= (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh} \text{ and} \\
\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eqn:t}
\end{align}
let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function. The RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects. The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.

@@

RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast, @salakhutdinov2012efficient, or @srivastava2013modeling for examples]. By stacking multiple layers of RBMs in a deep architecture, proponents of the models claim to produce the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp. 450]. The stacking is achieved by treating a hidden layer of one RBM as the visible layer in a second RBM, and so on, until the desired multi-layer architecture is created.

@@

### Near-degeneracy
In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q$ is a vector of statistics based on the adjacency matrix of a graph. Here, as earlier, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function. Let $C$ denote the convex hull of the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$. @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$. This makes sense because if a model is near-degenerate, in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ and can be expected to not be pulled deep into the interior of $C$.

A RBM model can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space of visibles and hiddens, $\mathcal{C}^{\nv + \nh}$. Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$. Here the mean of $\boldsymbol t(\boldsymbol x)$ is
$$
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
$$

@@

Considering a class of exponential families of distributions, @schweinberger2011instability introduced a concept of model defficiency related to *instability*. Instability can be roughly thought of as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$. To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how RBM models might be expanded to incorporate more and more visibles. When increasing the size of RBM models, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used). To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $V$ of visibles $(V_1, \dots, V_V)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{\nv}$ as
\begin{align}
\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) \label{eqn:elpr}
\end{align}
where $P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{\{h_j\}\in \mathcal{C}}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_\nv)$ for the visible variables $(V_1, \dots, V_\nv)$ under parameter vector $\boldsymbol \theta_\nv$.

In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio (\ref{eqn:elpr}) needs to stay bounded for the sequence of RBM models to be stable. That is, we make the following convention.

\begin{defn}[S-unstable RBM]
Let $\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh +  \nh*\nv}, \nv \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $\nh \equiv \nh(\nv) \ge 1$, can be an arbitrary function of the number $\nv$ of visibles. A RBM model formulation is \emph{Scheinberger-unstable} or \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) = \infty.
\end{align*}
\end{defn}
In other words, given any $C > 0$, there exists an integer $n_C > 0$ so that $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > C$ for all $\nv \ge n_C$.

@@

S-unstable RBM model sequences are undesirable for several reasons. One is that, as mentioned above, small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
$$
\Delta(\boldsymbol \theta_\nv) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{P_{\boldsymbol \theta_\nv}(v_1^*, \dots, v_\nv^*)} : (v_1, \dots, v_\nv) \& (v_1^*, \dots, v_\nv^*) \in \mathcal{C}^\nv \text{ differ by exactly one component}\right\},
$$
denoting the biggest log-probability ratio for a one component change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_\nv$. We then have the following result.

@@

In other words, if the probability ratio (\ref{eqn:elpr}) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability).

Additionally, S-unstable RBM model sequences are connected to the near-degeneracy of section \ref{near-degeneracy} (in which model sequences place all probability on a small portion of their sample spaces). To see this, define a modal set of possible outcomes (i.e. set of highest probability outcomes) in RBM models with parameters $\boldsymbol \theta_\nv, \nv \ge 1$ as
$$
M_{\epsilon, \boldsymbol \theta_\nv} \equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v) > (1-\epsilon)\max\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
$$
for a given $0 < \epsilon < 1$. Then S-unstable model sequences are guaranteed to be degenerate, as the following result shows.

@@

### Uninterpretability
For spatial Markov models, @kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by dependence parameters in a model being so extreme that marginal mean-structures fail to hold as anticipated by a model statement. We adapt this notion to RBM models as follows. Note that in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_\nv}$ and $\theta_{h_1}, \dots, \theta_{h_\nh}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_\nv, H_1, \dots, H_\nh$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh$. Hence, these main effect parameters have a clear interpretation under an independence model (one with $\theta_{ij} = 0$) but this interpretation (as well as an understanding of the nature of dependence) can break down as interaction parameters increase in magnitude relative to the size of the main effects. In such cases, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the models (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (destroying the model interpretation of dependence as local conditional modifications of an overall marginal mean structure). To assess which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ given independence where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$. Hence, $\boldsymbol \theta^*$ and $\boldsymbol \theta$ have the same main effects but $\boldsymbol \theta^*$ has $0$ dependence parameters. Uninterpretability is then avoided at a parametric specification $\boldsymbol \theta$ if the model expected value at $\boldsymbol \theta$ is not very different from the corresponding model expectation under independence. Using this, it is possible to investigate what parametric conditions lead to uninterpretability in a model versus those that guarantee interpretable models. If $||\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]||$ is large, then the RBM model with parameter vector $\boldsymbol \theta$ is said to be uninterpetable. The quantities to compare in the RBM case are
